---
output:
  html_document: default
  pdf_document: default
---
Red Wine Quality : An Exploratory Data Analysis by Nupur Shukla
===============================================================================

### Introduction

The objective of this analyis is to use Exploratory Data Analysis techniques to 
find "Which chemical properties influence the quality of wine ?" Being a wine 
enthusiast myself, I have experienced the price of wine largely depends on it's 
propularity otr appreciation in the social circle.I chose this dataset for 
analysis because I am interested to see whether human tasting can be related to 
the chemical properties of wine. This analysis can be used to further build a 
predictive model which can provide guidance to vineyards regarding the price and 
quality of of their wines. 

**The Data Set**

This tidy data set contains 1,599 red wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine,
providing a rating between 0 (very bad) and 10 (very excellent).

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(psych)
library(gridExtra)
library(ggcorrplot)
library(memisc)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
red_wine<-read.csv("wineQualityReds.csv")

```

**Summary of the Data Set**

```{r echo=FALSE,Summary_of_data}
names(red_wine)
str(red_wine)
summary(subset(red_wine,select=-X))

```


**Observations**

For most the input variables( all excluding 'quality') the mean is greater than 
the median.This indicates there maybe outliers in the dataset. It is interesting
to see some wine have no citric acid, which are often added for  'freshness' and 
flavor.

# Univariate Plots Section

**Historams showing variable distribution** 

Next, I will look at the univariate analysis and investigate variables 
individually to look for unusual data and patterns for futher analysis.

```{r echo=FALSE,summary_of_quality}
summary(red_wine$quality)
```

```{r echo=FALSE, Univariate_Plots_quality}
ggplot(aes(quality),data=red_wine)+
  geom_histogram(binwidth =1,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(3,9,1)) +
  ggtitle("Histogram for quality of wine")+
  theme(plot.title=element_text(hjust=0.5))
```

The distribution for Quality appears to be unimodal and normal with  most values 
concentrated in categories 5 and 6. The wine scores are in the range [3,8].

```{r echo=FALSE}
p1<-ggplot(aes(fixed.acidity),data=red_wine)+
  geom_histogram(binwidth =0.5,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(4,16,1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Fixed Acidity")

  
p2<-ggplot(aes(volatile.acidity),data=red_wine)+
  geom_histogram(binwidth =0.1,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(0,2.0,0.1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Volatile Acidity")
 

p3<-ggplot(aes(citric.acid),data=red_wine)+
  geom_histogram(binwidth =0.02,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(0,1,0.1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Citric Acid")
 

p4<-ggplot(aes(density),data=red_wine)+
  geom_histogram(binwidth =0.001,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(0,1.1,0.002)) +
  theme(axis.title.y=element_blank()) +
  xlab("Density")

p5<-ggplot(aes(pH),data=red_wine)+
  geom_histogram(binwidth =0.02,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(2.7,4.1,0.1)) +
  theme(axis.title.y=element_blank()) +
  xlab("pH")
  

p6<-ggplot(aes(alcohol),data=red_wine)+
  geom_histogram(binwidth=0.5,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(8,15,1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Alcohol")
  

grid.arrange(p1,p2,p3,p4,p5,p6,nrow=3,ncol=2,
             top="Histogram to show distributionfor the variable values",
             left="Number of Samples ")
```

```{r echo=FALSE,Univariate_Plots2}
p1<-ggplot(aes(residual.sugar),data=red_wine)+
  geom_histogram(binwidth =0.2,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(0,16,1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Residual Sugar")

  
p2<-ggplot(aes(chlorides),data=red_wine)+
  geom_histogram(binwidth =0.01,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(0,0.7,0.1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Chlorides")
 

p3<-ggplot(aes(free.sulfur.dioxide),data=red_wine)+
  geom_histogram(binwidth =1,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(1,72,10)) +
  theme(axis.title.y=element_blank()) +
  xlab("Free Sulfur DiOxide")
 

p4<-ggplot(aes(total.sulfur.dioxide),data=red_wine)+
  geom_histogram(binwidth =5,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(6,289,25)) +
  theme(axis.title.y=element_blank()) +
  xlab("Total Sulfur diOxide")

p5<-ggplot(aes(sulphates),data=red_wine)+
  geom_histogram(binwidth =0.02,col="black",fill="#99CCCC")+
  scale_x_continuous(breaks=seq(0.3,2,0.1)) +
  theme(axis.title.y=element_blank()) +
  xlab("Sulphates")
  



grid.arrange(p1,p2,p3,p4,p5,nrow=3,ncol=2,top="Histogram to show distribution
             for the variable values", left="Number of Samples ")

```


**Boxplots for each of the variables as another indicator of spread**

```{r echo=FALSE, Box_plots1}
m1<-ggplot(red_wine,aes(x="fixed acidity", y=fixed.acidity))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m2<-ggplot(red_wine,aes(x="volatile acidity", y=volatile.acidity))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m3<-ggplot(red_wine,aes(x="Citric Acid", y=citric.acid))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m4<-ggplot(red_wine,aes(x="density", y=density))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m5<-ggplot(red_wine,aes(x="pH", y=pH))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

grid.arrange(m1,m2,m3,m4,m5,nrow=1,ncol=5)

```

```{r echo=FALSE, Box_plots2}
m1<-ggplot(red_wine,aes(x="Alcohol", y=alcohol))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m2<-ggplot(red_wine,aes(x="Residual sugar", y=residual.sugar))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m3<-ggplot(red_wine,aes(x="sulphates", y=sulphates))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m4<-ggplot(red_wine,aes(x="free sulfur dioxide", y=free.sulfur.dioxide))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m5<-ggplot(red_wine,aes(x="total sulfur dioxide", y=total.sulfur.dioxide))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())

m6<-ggplot(red_wine,aes(x="chlorides", y=chlorides))+
  geom_boxplot(outlier.colour="red",outlier.shape=1) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank())


grid.arrange(m1,m2,m3,m4,m5,m6,nrow=1,ncol=6)

```

**Observations from the plots above :**

* All variables have outliers and most outliers are on the larger side.  
* Residual Sugar, Chlorides, Free Sulfur Dioxide, Total Sulfur Dioxide  
and Sulphates have a positively skewed distribution with quite a few outliers 
the larger side.Using a log transformation of these variables may help to fit 
these distributiontions into a more normal model or bell curve.
* Fixed Acidity follows a normal distribution in the range [4.6 g/dm3, 15.9 
g/dm3]. 50% of the data is between 4.6 g/dm3 and  7.90 g/dm3.
* _Volatile Acidity_ has a slighly right skewed normal distribution with a tail 
on the right side of the plot. The mean and median values are almost equal(0.52
g/dm3).
* Citric Acid has it's median at 0.26 and mean at 0.271.There are two main  
peaks in the plot between [0,0.02] and [0.48,0.5].According to the histogram, 
maximum number of wines do not contain citric acid. The Data requires 
transformation to better understand the distribution.
* Alcohol has a strong right skew but less outliers compared to other right 
skewed variables.
* pH and Density have a fairly normal distribution.

**Transforming data**

In order to reduce the skewness of variables; Residual Sugar, Chlorides, 
Free Sulfur Dioxide, Total Sulfur Dioxide and Sulphates  a log 
transformation is applied to make this distribution more "Normal". Skew data can 
be transformed to have a more normal distribution (ideal for linear or logistic 
regression  model )  using log or square root transformation techniques.
The histogram for Citirc Acid shows the distribution is 
not normal, a log transformation resulted in a left skewed distribution for this 
variable so I have applied a square root transformation to get a more normal 
looking plot except the spike at 0. The plots can be found below :

```{r echo=FALSE , Transformation_of_skewed_data}
p1<-ggplot(aes(residual.sugar),data=red_wine)+
  geom_histogram(binwidth =0.1,col="black",fill="#99CCCC")+
  scale_x_log10(breaks=c(1,2,5,10,20)) +
  theme(axis.title.y=element_blank()) 

  
p2<-ggplot(aes(chlorides),data=red_wine)+
  geom_histogram(binwidth =0.1,col="black",fill="#99CCCC")+
  scale_x_log10(breaks=c(0.01,0.03,0.08,0.2,0.4,0.7)) +
  theme(axis.title.y=element_blank()) 
 

p3<-ggplot(aes(free.sulfur.dioxide),data=red_wine)+
  geom_histogram(binwidth =0.1,col="black",fill="#99CCCC")+
  scale_x_log10(breaks=c(1,2,5,10,20,50,80)) +
  theme(axis.title.y=element_blank()) 
 

p4<-ggplot(aes(total.sulfur.dioxide),data=red_wine)+
  geom_histogram(binwidth =0.1,col="black",fill="#99CCCC")+
  scale_x_log10(breaks=c(1,5,10,20,50,80,150,300)) +
  theme(axis.title.y=element_blank()) 
  

p5<-ggplot(aes(sulphates),data=red_wine)+
  geom_histogram(binwidth =0.05,col="black",fill="#99CCCC")+
  scale_x_log10(breaks=c(0.2,0.4,0.6,0.8,1,1.4,2)) +
  theme(axis.title.y=element_blank()) 

p6<-ggplot(aes(citric.acid),data=red_wine)+
  geom_histogram(binwidth =0.05,col="black",fill="#99CCCC")+
  scale_x_sqrt(breaks=c(0,0.1,0.3,0.6,0.9)) +
  theme(axis.title.y=element_blank()) 


grid.arrange(p1,p2,p3,p4,p5,p6,nrow=2,ncol=3,
top="Histogram to show distribution for the transformed variable values",
left="Number of Samples ")

```

Internet articles suggest a good sign of quality is balance.A wine is balanced 
when all of the wine's components (e.g. sweetness, acidity, tannins, alcohol) 
blend together. The balance or potential to be balanced after ageing is a sign 
of quality.
Next, Let's create a categorical variable to distinguish 'loW'(score<5), 
'medium'(5=<score<7) and 'high' (score>=8) quality wines.

```{r, adding_variable}
# Adding a new feature wine_category which distinguishes wines into 3 categories
red_wine$wine_category<-0
red_wine$wine_category<-ifelse(red_wine$quality<5,'low',red_wine$wine_category)
red_wine$wine_category<-ifelse((red_wine$quality>=5 & red_wine$quality<7) ,
                               'medium',red_wine$wine_category)
red_wine$wine_category<-ifelse(red_wine$quality>=7 ,'high',
                               red_wine$wine_category)
red_wine$wine_category<-ordered(red_wine$wine_category,levels=c('low','medium',
                                                                'high'))
```

```{r echo=FALSE}
ggplot(aes(wine_category),data=red_wine)+
  geom_bar(col="black",fill="#99CCCC")+
  ggtitle("Barplot showing counts for different wine categories")+
  theme(plot.title=element_text(hjust=0.5))
```

The Barplot above clearly indicates that maximum samples in this data set are
of medium quality. 
Since, quality is indicating the wine category I will now convert quality into a 
factor variable.
```{r}
red_wine$quality_factor<-factor(red_wine$quality)
```

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 red wine in this data set with 12 features (fixed.acidity,
volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, 
total.sulfur.dioxide, density, pH, sulphates and quality).The Univariate plots 
above show most features observe a normal distribution  but some outliers

### What is/are the main feature(s) of interest in your dataset?

The focus of this analysis is to identify the features that influence the  
quality of wine which makes it the most important feature in this data set.
However, a good quality wine has a balance of sweetness( residual sugar ), 
acidity( fixed acidity + volatile acidity) and alcohol( alcohol) level. 
I would like to find out which features are strongly correlated with quality in 
further sections.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Other than features mentioned previously, density (dependent on alcohol and 
sugar content) and sulfur dioxide(both free and total) are likely to 
contribute to the quality of wine.

### Did you create any new variables from existing variables in the dataset?

wine_category a categorical variable to distinguish 'low'(score<5), 'medium'
(5=<score<7) and 'high' (score>=8) quality wines. quality_factor as another
categorical variable.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Yes,in order to reduce the skewness of variables; Residual Sugar, Chlorides, 
Free Sulfur Dioxide, Total Sulfur Dioxide, Sulphates and Citric Acid a 
log /square root transformation is applied to obtain a normal or close to
normal distribution.

# Bivariate Plots Section

**Correlation Matrix**

Next, we look at the correlation matrix for the datasset  to find relationships 
between features.

```{r echo=FALSE , correlation_matrix}
corrmat<-cor(x=red_wine[,2:13])
corrmat
```
```{r echo=FALSE, correlation_plot}
ggcorrplot(corrmat , method="circle")
```

The Correlation Matrix output suggests the following:

* Alcoholshows a strong positive correlation with quality followed by 
sulphates and citric Acid .
* Volatile acidity responsible for unpleasant smell in wines has a strong
negative correlation with quality. It also shows a negative correlation with 
citric Acid.
* Fixed Acidity shows a positive correlation with citric acid (fixed acid 
component) and density. It also shows a negative correlation with pH and
Volatile acidity. 
* Free sulfur dioxide shows a strong positive correlation with total sulfur di
oxide. 

Below are some plots summarising the relationships between features:

** Quality vs Alcohol **

```{r echo=FALSE, scatter_quality_vs_alcohol}
ggplot(red_wine, aes(x=quality_factor,y=alcohol, color=quality_factor))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method = 'lm')+
  ggtitle("Scatterplot for Alcohol vs Quality")+
  theme(plot.title=element_text(hjust=0.5))+
  ylab("Alcohol(%)")+
  xlab("Quality(score)")

```

The plot above clearly shows a positive relationship between alcohol(% by volume
)and  quality score. 

**Summary of alcohol content grouped by wine category**

```{r echo=FALSE, alcohol_quality_summmary}
red_wine %>%
  group_by(wine_category) %>%
  summarise(mean_alcohol=mean(alcohol),
           median_alcohol=median(alcohol),
           max_alcohol=max(alcohol))

```

```{r echo=FALSE, box_quality_vs_alcohol}
ggplot(red_wine,aes(x=wine_category, y=alcohol))+
  geom_jitter( alpha = 0.3)+
  geom_boxplot(alpha=0.5,colour="blue",outlier.shape=1)+
  stat_summary(fun.y="mean",geom="point",color="red",
               shape=8,size=4)+
  ggtitle("Boxplot for Wine category based on Quality score vs Alcohol")+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Wine Category based on Quality score")+
  ylab("Alcohol(%)")
```

_Alcohol_ content does not show an strong significance in quality prediction.
Both low and medium quality wines have 50% of samples containing more than 10% 
by volume alcohol content. Box plot for medium  quality wines show some outliers
(high alcohol content) with maximum content of about 14.9% by vol. which is 
greater than the maximum alcohol content found in high quality wines of 14% by
vol. However, for both low and medium quality wines the median is more samples
have low alcohol content whereas fro high quality wines more sameples have high 
alcohol content.

```{r echo=FALSE, scatter_volatileacidity_vs_quality}
ggplot(red_wine, aes(x=quality_factor,y=volatile.acidity, color=quality_factor))+
  geom_jitter(alpha=1/3)+
  ggtitle("Scatterplot for Volatile Acidity vs Quality")+
  theme(plot.title=element_text(hjust=0.5))+
  ylab("Volatile Acidity")+
  xlab("Quality(score)")

```

Plot above shows a negative correlation between volatile acidity and wine 
quality.

```{r, volatile_Acidity_quality_summary}
red_wine %>%
  group_by(wine_category) %>%
  summarise(mean_volatile_acidity=mean(volatile.acidity),
           median_volatile_acidity=median(volatile.acidity),
           sd_volatile_acidity=sd(volatile.acidity))

```

```{r echo=FALSE, Boxplot_volatileacidity_quality}

ggplot(red_wine,aes(x=wine_category, y=volatile.acidity))+
  geom_boxplot(outlier.colour="red",outlier.shape=1)+
  geom_jitter( alpha = 0.3)+
  geom_boxplot(alpha=0.5,colour="blue",outlier.shape=1)+
  stat_summary(fun.y="mean",geom="point",color="red",
               shape=8,size=4)+
  ggtitle("Boxplot for Wine category based on Quality score vs Volatile Acidity")+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Wine Category based on Quality score")+
  ylab("Volatile Acidity")
```

We can clearly see a the quality increases (from low to high) the mean, median 
and standard deviation for volatile acidity decreases. Even though more samples
for medium quality wines contain high volatile acidity the median value (0.58)
is less than the median value for low quality wines(o.68). It is interesting to 
see that the spread of volatile acidity is right skewed i.e. for most samples 
in low quality wines the voltile acid content is low and for high quality wines
the spread is left skewed and most samples have high volatile acidity content.
Getting more samples of low ad high quality scores will help understand
relationships between features much better.,

```{r echo=FALSE , scatterplot_citric_acid_quality1}
ggplot(red_wine, aes(x=quality_factor,y=citric.acid, color=quality_factor))+
  geom_jitter(alpha=1/3)+
  ggtitle("Scatterplot for Citric Acid vs Quality")+
  theme(plot.title=element_text(hjust=0.5))+
  ylab("Citric Acid")+
  xlab("Quality(score)")

```
```{r echo= FALSE, boxplot_citricacid_vs_quality}
ggplot(red_wine,aes(x=wine_category, y=citric.acid))+
  geom_boxplot(outlier.colour="red",outlier.shape=1)+
  geom_jitter( alpha = 0.3)+
  geom_boxplot(alpha=0.5,colour="blue",outlier.shape=1)+
  stat_summary(fun.y="mean",geom="point",color="red",
               shape=8,size=4)+
  ggtitle("Boxplot for wine Category vs Citric Acid content")+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Wine category based on quality score")+
  ylab("Citric Acid content")
```

The scatter plot and box plot for citirc acid vs quality show positive 
correlation between the features. The box plot for citric acid
shows very less outliers compared to  other plots and might be used for 
prediction of quality.

```{r echo=FALSE, scatterplot_sulphates_vs_quality}
ggplot(red_wine, aes(x=quality_factor,y=sulphates, color=quality_factor))+
  geom_jitter(alpha=1/3)+
  ggtitle("Scatterplot for sulphates vs Quality")+
  theme(plot.title=element_text(hjust=0.5))+
  ylab("Sulphates")+
  xlab("Quality(score)")


```
```{r echo=FALSE, boxplot_sulphates_vs_quality}
ggplot(red_wine,aes(x=wine_category, y=sulphates))+
  geom_jitter( alpha = 0.3)+
  geom_boxplot(alpha=0.5,colour="blue",outlier.shape=1)+
  stat_summary(fun.y="mean",geom="point",color="red",
               shape=8,size=4)+
  ggtitle("Boxplot for wine Category vs Sulphates")+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Wine category based on quality score")+
  ylab("Sulphates")
```

The plots for sulphates vs quality show a positive relationship but most 
outliers are in medium quality wines.
Next, the plots below summarize relationships betweeen features other than 
quality.

```{r echo=FALSE, scatterplot_other_features}
p1<-ggplot(red_wine, aes(x=citric.acid,y=volatile.acidity))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method='lm')+
 # ggtitle("Scatterplot for citric acid vs volatile acidity")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("citric Acid ")+
  ylab("Volatile Acidity")
p2<-ggplot(red_wine, aes(x=alcohol,y=density))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method='lm')+
 # ggtitle("Scatterplot for alcohol vs density")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("alcohol")+
  ylab("density")

p3<-ggplot(red_wine, aes(x=fixed.acidity,y=density))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method='lm')+
 # ggtitle("Scatterplot for Fixed Acidity vs density")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Fixed acidity")+
  ylab("density")

p4<-ggplot(red_wine, aes(x=fixed.acidity,y=citric.acid))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method='lm')+
  #ggtitle("Scatterplot for Fixed Acidity vs Citric Acid")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Fixed acidity")+
  ylab("Citric Acid")

p5<-ggplot(red_wine, aes(x=fixed.acidity,y=pH))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method='lm')+
  #ggtitle("Scatterplot for Fixed Acidity vs pH")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Fixed acidity")+
  ylab("pH")

p6<-ggplot(red_wine, aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide))+
  geom_jitter(alpha=1/3)+
  geom_smooth(method='lm')+
  #ggtitle("Scatterplot for Free sulfurdioxide vs Total sulfurdioxide")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Free sulfurdioxide")+
  ylab("Total sulfurdioxide")

grid.arrange(p1,p2,p3,p4,p5,p6,nrow=3,ncol=2)

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

*Alcohol content for high quality wines seems to be high but does not have any
significant bias.
*Density and alcohol content have a negative relationship. Samples with less
density have more alcohol content.
*Fixed acidity is storngly correlated with pH, density and Citric Acid.
*Wine samples with high sulphates have higher quality. Sulphates show a positive
correlation with quality but there are significant number of outliers.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Residual sugar, chlorides and sulphates do not show any significant relationships
with any other feature. Some of the strongest correlations are between chemical 
attribute variables, and not the main feature of interest(quality). There is
a strong positive relationship between _fixed acidity_ and _density_, 
_free sulfur dioxide_ and _total sulfur dioxide_, _citric acid_ and 
_fixed acidity_. _Volatile acidity_ and _citric acid_, _fixed acidity_ and _pH_,
_alcohol_ and _density_ show strong negative relationships.

### What was the strongest relationship you found?

_Alcohol_ shows the strongest association with feature of interest, _quality_ 
(r-squared=0.476).Among other features _pH_ and _fixed acidity_ show highest 
correlation (r-squared=-0.683)

# Multivariate Plots Section

```{r echo=FALSE, alcohol_vs_citric_quality}
ggplot(red_wine, aes(x=alcohol,y=citric.acid,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$alcohol, .01), 
             quantile(red_wine$alcohol, .99)),
    ylim = c(quantile(red_wine$citric.acid, .01), 
             quantile(red_wine$citric.acid, .99))
    )+
  theme_dark()+
  geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Alcohol vs Citric Acid vs Quality")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Alcohol ")+
  ylab("Citric Acid")



```

We can see in the above plot that high quality wines are spreadout in the region 
where alcohol>=10 and samples become more concentrated when citric acid content 
is more that 0.3 (Approx.).

```{r echo=FALSE, alcohol_vs_volatile_quality}
ggplot(red_wine, aes(x=alcohol,y=volatile.acidity,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$alcohol, .01), 
             quantile(red_wine$alcohol, .99)),
    ylim = c(quantile(red_wine$volatile.acidity, .01), 
             quantile(red_wine$volatile.acidity, .99))
    )+
  theme_dark()+
  geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Alcohol vs Volatile acidity vs Quality")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Alcohol")+
  ylab("Volatile Acidity")

```

_Alcohol_ and _Volatile acidity_ have the strongest positive and negative 
correlation with quality, respectively. There is a fairly clear separation 
between the low quality wines(upper left) and high quality wines(lower right) 
but the medium quality wines are spreadout throughout. Hence, a model with these
two features alone will not be a good predictor of wine quality.

```{r echO=FALSE, volatile_citric_quality_}
ggplot(red_wine, aes(x=volatile.acidity,y=citric.acid,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$volatile.acidity, .01), 
             quantile(red_wine$volatile.acidity, .99)),
    ylim = c(quantile(red_wine$citric.acid, .01), 
             quantile(red_wine$citric.acid, .99))
    )+
  theme_dark()+
   geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Volatile acidity vs Citric Acid vs Quality")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  ylab("Citric Acid")+
  xlab("Volatile Acidity")


```

In the graph above, low quality wines have lower values for citric acid and 
higher volatile acidity, and high quality wines have high ctric acid values but
low volatile acidity.

```{r echo=FALSE}
ggplot(red_wine, aes(x=fixed.acidity,y=density,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$fixed.acidity, .01), 
             quantile(red_wine$fixed.acidity, .99)),
    ylim = c(quantile(red_wine$density, .01), 
             quantile(red_wine$density, .99))
    )+
  theme_dark()+
  geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Fixed acidity vs Density vs Quality")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  ylab("Density")+
  xlab("Fixed Acidity")
```

This graph reveals an interesting pattern where low and medium quality wines
fall more towards the upper left side of the graph while high quality wines fall
more towards the lower right side of the graph. Higher quality wines will have 
lower density value.

```{r echo=FALSE}
ggplot(red_wine, aes(x=alcohol,y=density,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$alcohol, .01), 
             quantile(red_wine$alcohol, .99)),
    ylim = c(quantile(red_wine$density, .01), 
             quantile(red_wine$density, .99))
    )+
  theme_dark()+
  geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Alcohol vs Density vs Quality")+
  theme(plot.title=element_text(hjust=0.5))+
  ylab("Density")+
  xlab("Alcohol")
```

As discovered previously high quality wines have more alcohol content and lower
density values. The graph above confirms this trend in the features.

**Regression Model**

I will now create a linear regression model to predict wine quality.All variables
are added to the model in progression( according to their relationship with 
quality).

```{r , Regression_Model}
m1<-lm(data=red_wine,quality~volatile.acidity)
m2<-update(m1,~.+alcohol)
m3<-update(m2,~.+sulphates)
m4<-update(m3,~.+citric.acid)
m5<-update(m4,~.+chlorides+total.sulfur.dioxide+density+fixed.acidity)
mtable(m1,m2,m3,m4,m5)

```

The r-squared value improved from 0.153(m1) to 0.356(m5) with addtion of 
variables. _Alcohol_, _volatile acidity_ and _sulphates_ show highest correlation
with quality but the r-squared value is very low, 0.336(m3). Also, addition of
_citric acid_ (m4) did not change the r-squared value at all.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

High quality wine samples are more concentrated when _citric acid_ is more than 
0.3 g/dm^3 and _alcohol_ more than 10.5. Fixed acidity and density may strengthen
 predicion model for quality as a clear pattern in wine samples was observed in
 the scatterplot.

### Were there any interesting or surprising interactions between features?

Although _alcohol_ and _volatile acidity_  do not have a very strong relationship
with each other but when plotted together, a clear separation between low and 
high quality wines can be seen. 


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created a linear regression model adding features one by  one as predictor 
variables and wine quality as the outcome variable. The overall r-squared value
is 0.356 which is quite low. _alcohol_ is  strongly related to quality and shows
a substantial improvement in r-squared value form 0.153 to 0.317 when added to 
the model.This makes it the most important predictor variable in my model. The 
strength of this model is it does not use many predictor variables that are 
weakly correlated with quality. The model is limited by the lack of data for
low(0,4] and high(6,10] quality wine samples. Most of the ouliers are 
concentrated in the medium quality wine bracket which makes the model weak.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(red_wine,aes(x=wine_category, y=alcohol))+
  geom_jitter( alpha = 0.3)+
  geom_boxplot(alpha=0.5,colour="blue",outlier.shape=1)+
  stat_summary(fun.y="mean",geom="point",color="red",
               shape=8,size=4)+
  ggtitle("Boxplot for Wine category based on Quality score vs Alcohol")+
  theme(plot.title=element_text(hjust=0.5))+
  xlab("Wine Category based on Quality score")+
  ylab("Alcohol(%)")
```

### Description One

_Alcohol_ has the strongest relationship with _quality_ (r-squared=0.476) and 
also, the most important feature in the predictor model. High Quality wines have
high alcohol content compared to low quality wines.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(red_wine, aes(x=fixed.acidity,y=density,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$fixed.acidity, .01), 
             quantile(red_wine$fixed.acidity, .99)),
    ylim = c(quantile(red_wine$density, .01), 
             quantile(red_wine$density, .99))
    )+
  theme_dark()+
  geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Fixed acidity vs Density vs Quality")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  ylab("Density (g/cm^3)")+
  xlab("Fixed Acidity (Tartaric Acid-g/dm^3)")
```

### Description Two

_Fixed Acidity_ and _density_ have a strong correlation(r-squared=0.668). The
plot shows an interesting trend where low and medium quality wines fall more 
towards the upper left side of the graph and high quality wines are towards
the lower right side of the graph.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(red_wine, aes(x=alcohol,y=volatile.acidity,color=wine_category))+
  geom_jitter(alpha=0.6,size=2)+
  coord_cartesian(
    xlim = c(quantile(red_wine$alcohol, .01), 
             quantile(red_wine$alcohol, .99)),
    ylim = c(quantile(red_wine$volatile.acidity, .01), 
             quantile(red_wine$volatile.acidity, .99))
    )+
  theme_dark()+
  geom_smooth(method="lm",se=FALSE,size=1)+
  scale_color_brewer(palette="Reds",guide=guide_legend(title='wine category'))+
  ggtitle("Scatterplot for Alcohol vs Volatile acidity vs Quality")+
  theme(plot.title=element_text(hjust=0.5,size=12))+
  xlab("Alcohol(% by volume)")+
  ylab("Volatile Acidity (Acetic Acid- g/dm^3)")

```

### Description Three

These two chemical attributes are the most important predictors of wine quality 
in the data set. It clearly shows that high and low quality wines are at 
opposite ends of the graph but medium quality wines are spread throughout.Higher 
quality wines tend to have lower volatile acidity values and higher alcohol 
values and lower quality wines show the opposite pattern.
------

# Reflection

This dataset contained 1,599 observations of red wines. Each wine received a 
quality rating and had information on 11 different chemical attributes. Most
of the analysis in this project was done on the  4 feautres ( _alcohol(0.476)_,
_volatile acidity(-0.39)_, _sulphates(0.251)_, _citric acid(0.226)_) that have 
the highest correlation with _quality_. Apart form these, some features showed
interesting relationships and patterns in wine quality  when plotted against 
each other for example, _Density_ vs _Fixed acidity_ (plot 2)  and _Volatile_
_acidity_ vs _alcohol_(plot 3).

In the end , I created a regression model progressively adding the top 4 
features mentioned above. _Alcohol_, _volatile acidity_ and _sulphates_ show 
highest correlation with quality but this model(m3) is able to explain only 
33.6% of the existing cases. Also, addition of _citric acid_ (m4) did not change 
the r-squared value at all. However, adding some variables with some correlation 
with quality improved the value t0 35.6%.

One of the major challenges in this analysis was the limitations of the dataset. 
The sign of quality in wines is when all components ( sweetness,acidity,tannins)
blend together.*Maturity*, *colour* (depending on different 
grape varities), *climatic conditions* (The higher the level of sugar in the 
grapes before fermentation, the higher potential alcohol the wine will have, 
i.e. hotter countries tend to produce wines higher in alcohol), *viticulture*
(way grapes are grown) and *vinification* (winemaking) are some important 
factors that effect the quality of wine and cannot be captured by chemical
attribute variables.Also,Current data consists of samples collected from a s
pecific portugal region.It will be intresting to obtain datasets across various 
wine making regions to eliminate any bias created by any secific qualities of 
the product.

Further analyis can be useful in creating a more sophisticated prediction model
for wine quality if mmore data of extreme quality cases (both high and 
low-quality) along with non-chemical attributes can be collected. 